The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-08-23 Independence_CH/forcing.tex
Locales used for forcing

Further locales gather the assumption of transitivity of $M$ and
particular replacement instances expressed by means of the
$\isatt{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate ($@$ denotes
list concatenation):
\begin{multline}\label{eq:replacement_assm_def}
\varphi \in \formula  \limp \mathit{env} \in \isatt{list}(M) \limp \isatt{arity}(\varphi) \leq 2+ \isatt{length}(\mathit{env}) \limp \\
 \isatt{strong{\uscore}replacement}(\isatt{\#\#} M, \lambda x\, y.\ (M, [x,y]
\mathbin{@} \mathit{env}  \models \varphi))
\end{multline}
%% % Same as above
%% \begin{isabelle}
%% {\isasymphi}\;{\isasymin}\;formula\ {\isasymlongrightarrow}\ env\;{\isasymin}\;list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow} arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\;{\isasymle}\;{\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
%% \ \ \ \ strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x\ y{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M\ {\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}y{\isacharbrackright}{\kern0pt}{\isacharat}{\kern0pt}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}
%% \end{isabelle}
In turn, the \isatt{{\isasymcdot}Replacement{\isasymcdot}} function
takes a formula code and returns the corresponding replacement
instance:
\begin{isabelle}
{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\isanewline
{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}env{\isachardot}{\kern0pt}\ replacement{\isacharunderscore}{\kern0pt}assm{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ env{\isacharcomma}{\kern0pt}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}
\end{isabelle}
%
%% Some of those instances, which we included for simplicity of design,
%% are actually “fake” instances in that they can be obtained from
%% Powerset, or from other replacement instances already available in the
%% respective context. An example of this is the sole instance of the
%% \locale{M{\uscore}basic} locale, originally from
%% \session{ZF-Constructible}; we managed to eliminate it but in
%% doing so we had to reprove many lemmas from
%% that session in the weakened context.

@@

\subsection{The fundamental theorems}\label{sec:fundamental-theorems}
In order to fix notation, let $\lb \PP, {\preceq} ,\1\rb \in M$ be a forcing notion. The
$\val$ interpretation function takes $3$ arguments so that if $G\sbq \PP$, we have
$M[G]\defi \{ \val(\PP,G,\punto{a}) : \punto{a}\in M \}$.

@@

  $\phi\in\formula$ and $\punto{a}_0,\dots,\punto{a}_n\in M$,

@@

  “$p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]$”

@@

  “$M, [p,\PP,\preceq,\1, \punto{a}_0,\dots,\punto{a}_n]  \models

@@

      \exists p\in G.\ \; p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]

@@

      M[G], [\val(\PP,G,\punto{a}_0),\dots,\val(\PP,G,\punto{a}_n)]

@@

  \item \label{item:density-lemma} (Density Lemma) $p \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]$

@@

    $\{q\in \PP :  q \forces \phi\ [\punto{a}_0,\dots,\punto{a}_n]\}$

@@

\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M{\isacharunderscore}{\kern0pt}generic{\isacharparenleft}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline

@@

\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\kern0pt}{\isasymexists}p{\isasymin}G{\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\kern0pt}\ \ \ {\isasymlongleftrightarrow}\ \ \ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ map{\isacharparenleft}{\kern0pt}val{\isacharparenleft}{\kern0pt}P{\isacharcomma}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isacharcomma}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}

@@

\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ \ \ {\isasymequiv}\ \ \ M{\isacharcomma}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}p{\isacharcomma}{\kern0pt}P{\isacharcomma}{\kern0pt}leq{\isacharcomma}{\kern0pt}{\isasymone}{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ forces{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline

>>>>

Further locales gather
particular replacement instances expressed by means of the
$\isatt{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate ($@$ denotes
list concatenation):
\begin{multline}\label{eq:replacement_assm_def}
\varphi \in \formula  \limp \mathit{env} \in \isatt{list}(M) \limp \isatt{arity}(\varphi) \leq 2+ \isatt{length}(\mathit{env}) \limp \\
 \isatt{strong{\uscore}replacement}(\isatt{\#\#} M, \lambda x\, y.\ (M, [x,y]
\mathbin{@} \mathit{env}  \models \varphi))
\end{multline}
%% % Same as above
%% \begin{isabelle}
%% {\isasymphi}\;{\isasymin}\;formula\ {\isasymlongrightarrow}\ env\;{\isasymin}\;list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow} arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}\;{\isasymle}\;{\isadigit{2}}\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymlongrightarrow}\isanewline
%% \ \ \ \ strong{\isacharunderscore}{\kern0pt}replacement{\isacharparenleft}{\kern0pt}{\isacharhash}{\kern0pt}{\isacharhash}{\kern0pt}M{\isacharcomma}{\kern0pt}{\isasymlambda}x\ y{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}M\ {\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}x{\isacharcomma}{\kern0pt}y{\isacharbrackright}{\kern0pt}{\isacharat}{\kern0pt}env\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}
%% \end{isabelle}
In turn, the \isatt{{\isasymcdot}Replacement{\isasymcdot}} function
takes a formula code and returns the corresponding replacement
instance:
\begin{isabelle}
{\isasymphi}\ {\isasymin}\ formula\ {\isasymLongrightarrow}\isanewline
{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ {\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymcdot}Replacement{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymcdot}{\isacharparenright}{\kern0pt}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}{\kern0pt}{\isasymforall}env{\isachardot}{\kern0pt}\ replacement{\isacharunderscore}{\kern0pt}assm{\isacharparenleft}{\kern0pt}M{\isacharcomma}{\kern0pt}\ env{\isacharcomma}{\kern0pt}\ {\isasymphi}{\isacharparenright}{\kern0pt}{\isacharparenright}
\end{isabelle}
%
%% Some of those instances, which we included for simplicity of design,
%% are actually “fake” instances in that they can be obtained from
%% Powerset, or from other replacement instances already available in the
%% respective context. An example of this is the sole instance of the
%% \locale{M{\uscore}basic} locale, originally from
%% \session{ZF-Constructible}; we managed to eliminate it but in
%% doing so we had to reprove many lemmas from
%% that session in the weakened context.

@@

\subsection{The fundamental theorems}\label{sec:fundamental-theorems}
At this point, we work inside the locale \isatt{M{\uscore}ctm1} that
assumes $M$ to be
countable and transitive, and satisfies some fragment of
$\ZFC$.
\footnote{%
Namely, Zermelo set theory plus the replacement instances included in
\locale{M{\uscore}ZF1} and \locale{M{\uscore}ZF{\uscore}ground}.
}%
conjunction of the sets This is further extended by assuming a forcing notion
$\lb \PP, {\preceq} ,\1\rb \in M$. The actual implementation reads:

\begin{isabelle}
\isacommand{locale}\isamarkupfalse%
\ forcing{\isacharunderscore}{\kern0pt}notion\ {\isacharequal}{\kern0pt}\isanewline
\ \ \isakeyword{fixes}\ P\ {\isacharparenleft}{\kern0pt}{\isacartoucheopen}{\isasymbbbP}{\isacartoucheclose}{\isacharparenright}{\kern0pt}\ \isakeyword{and}\ leq\ \isakeyword{and}\ one\ {\isacharparenleft}{\kern0pt}{\isacartoucheopen}{\isasymone}{\isacartoucheclose}{\isacharparenright}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}P{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymone}\ {\isasymin}\ {\isasymbbbP}{\isachardoublequoteclose}\isanewline
\ \ \ \ \isakeyword{and}\ leq{\isacharunderscore}{\kern0pt}preord{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ {\isachardoublequoteopen}preorder{\isacharunderscore}{\kern0pt}on{\isacharparenleft}{\kern0pt}{\isasymbbbP}{\isacharcomma}{\kern0pt}leq{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \ \ \isakeyword{and}\ one{\isacharunderscore}{\kern0pt}max{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymforall}p{\isasymin}{\isasymbbbP}{\isachardot}{\kern0pt}\ {\isasymlangle}p{\isacharcomma}{\kern0pt}{\isasymone}{\isasymrangle}{\isasymin}leq{\isachardoublequoteclose}
\end{isabelle}

\begin{isabelle}
\isacommand{locale}\isamarkupfalse%
\ forcing{\isacharunderscore}{\kern0pt}data{\isadigit{1}}\ {\isacharequal}{\kern0pt}\ forcing{\isacharunderscore}{\kern0pt}notion\ {\isacharplus}{\kern0pt}\ M{\isacharunderscore}{\kern0pt}ctm{\isadigit{1}}\ {\isacharplus}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymbbbP}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline
\ \ \ \ \isakeyword{and}\ leq{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M{\isacharcolon}{\kern0pt}\ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}leq\ {\isasymin}\ M{\isachardoublequoteclose}
\end{isabelle}
In order to fix notation, the
$\val$ interpretation function takes as its first argument the filter
and then the name, so we have
$M[G]\defi \{ \val(G,\tau) : \tau\in M \}$.

@@

  $\phi\in\formula$ and $\tau_0,\dots,\tau_n\in M$,

@@

  “$p \forces \phi\ [\tau_0,\dots,\tau_n]$”

@@

  “$M, [p,\PP,\preceq,\1, \tau_0,\dots,\tau_n]  \models

@@

      \exists p\in G.\ \; p \forces \phi\ [\tau_0,\dots,\tau_n]

@@

      M[G], [\val(G,\tau_0),\dots,\val(G,\tau_n)]

@@

  \item \label{item:density-lemma} (Density Lemma) $p \forces \phi\ [\tau_0,\dots,\tau_n]$

@@

    $\{q\in \PP :  q \forces \phi\ [\tau_0,\dots,\tau_n]\}$

@@

\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\isanewline

@@

\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\kern0pt}{\isasymexists}p{\isasymin}G{\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\kern0pt}\ \ \ {\isasymlongleftrightarrow}\ \ \ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ map{\isacharparenleft}{\kern0pt}val{\isacharparenleft}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isacharcomma}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}

@@

\ \ {\isachardoublequoteopen}p\ {\isasymtturnstile}\ {\isasymphi}\ env\ \ \ {\isasymequiv}\ \ \ M{\isacharcomma}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isacharbrackleft}{\kern0pt}p{\isacharcomma}{\kern0pt}{\isasymbbbP}{\isacharcomma}{\kern0pt}leq{\isacharcomma}{\kern0pt}{\isasymone}{\isacharbrackright}{\kern0pt}\ {\isacharat}{\kern0pt}\ env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ forces{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}

### 2022-08-23 Independence_CH/sample_proof.tex
Locales used for forcing

\ {\isacharquery}{\kern0pt}Q{\isacharequal}{\kern0pt}{\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isachardoublequoteclose}

@@

\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline

@@

%% \ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}P\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ transitivity\isanewline

@@

\ {\isacharquery}{\kern0pt}{\isasymtheta}{\isacharequal}{\kern0pt}{\isachardoublequoteopen}{\isacharbraceleft}{\kern0pt}{\isasymlangle}{\isasymsigma}{\isacharcomma}{\kern0pt}p{\isasymrangle}\ {\isasymin}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}P\ {\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymcdot}{\isadigit{0}}\ {\isasymin}\ {\isadigit{1}}{\isasymcdot}\ {\isacharbrackleft}{\kern0pt}{\isasymsigma}{\isacharcomma}{\kern0pt}{\isasymchi}{\isacharbrackright}{\kern0pt}\ {\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}

>>>>

\ {\isacharquery}{\kern0pt}Q{\isacharequal}{\kern0pt}{\isachardoublequoteopen}Pow{\isacharparenleft}{\kern0pt}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}\isasymbbbP{\isacharparenright}{\kern0pt}\ {\isasyminter}\ M{\isachardoublequoteclose}

@@

\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}\isasymbbbP\ {\isasymin}\ M{\isachardoublequoteclose}\ {\isachardoublequoteopen}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}\ {\isasymin}\ M{\isachardoublequoteclose}\isanewline

@@

%% \ one{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}\isasymbbbP\ P{\isacharunderscore}{\kern0pt}in{\isacharunderscore}{\kern0pt}M\ transitivity\isanewline

@@

\ {\isacharquery}{\kern0pt}{\isasymtheta}{\isacharequal}{\kern0pt}{\isachardoublequoteopen}{\isacharbraceleft}{\kern0pt}{\isasymlangle}{\isasymsigma}{\isacharcomma}{\kern0pt}p{\isasymrangle}\ {\isasymin}domain{\isacharparenleft}{\kern0pt}{\isasymtau}{\isacharparenright}{\kern0pt}{\isasymtimes}\isasymbbbP\ {\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymcdot}{\isadigit{0}}\ {\isasymin}\ {\isadigit{1}}{\isasymcdot}\ {\isacharbrackleft}{\kern0pt}{\isasymsigma}{\isacharcomma}{\kern0pt}{\isasymchi}{\isacharbrackright}{\kern0pt}\ {\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}
